### 2022-06-02 chapters/fairness/fif/framework.tex
minor edits in many parts of the thesis

%in Section~\ref{??}, and conclude by extending influence computation to other group fairness metrics in Section~\ref{??}
%\subsection{An Axiomatic Formulation}
%\label{sec:problem_formulation}
\noindent\textbf{FIF: An Axiomatic Formulation.} We are given a binary classifier  $\alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} $, a dataset $ \mathbf{D} = \{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n $, and a  fairness metric $ f(\alg, \mathbf{D}) \in \mathbb{R}^{\geq 0} $. Our objective is to compute the influences of non-sensitive features on $ f $. We compute influence of each subset of non-sensitive features $ \mathbf{X}_{\mathbf{S}} $, where $ \mathbf{S} = \{S_i | 1 \le S_i \le \numnonsensitive\} \subseteq [\numnonsensitive] \setminus \emptyset $ is a non-empty subset of indices.

\begin{definition}[Fairness Influence Function]
Fairness Influence Function (FIF) $ w_{\mathbf{S}}:\mathbf{X}_{\mathbf{S}} \rightarrow \mathbb{R} $ measures the quantitative contribution of the subset of features $ \mathbf{X}_{\mathbf{S}} \subseteq \mathbf{X}_{[k]}$ on the incurred bias $ f(\alg, \mathbf{D}) $ of the classifier $ \alg $ for dataset $\mathbf{D}$. We refer $ w_{\mathbf{S}} $ as an \emph{individual influence} when $ |\mathbf{S}| = 1 $ and an \emph{inter-sectional influence} when $ |\mathbf{S}| > 1 $\textemdash particularly, a second-order influence when $ |\mathbf{S}| = 2 $ and a higher-order influence when $ 3 \le |\mathbf{S}| \le \numnonsensitive $. %\unsure{What do we mean by function here? Isn't FIF scalar?}
\end{definition}	
Since FIF is meant to attribute the bias of the classifier to the subset of features, it is natural to choose a function such that the FIFs of all the subsets of features exactly adds up to the total bias. This leads to the following additivity axiom.
%{\fairXplainer} is an axiomatic approach to relate individual and inter-sectional influences with associated fairness metrics, which we formalize in the following.
\begin{axiom}[Additivity of influence]
	\label{axm:additivity}
	Let $ f(\alg, \mathbf{D}) \in \mathbb{R}^{\geq 0}$ be the bias of the classifer and $ w_{\mathbf{S}}:\mathbf{X}_{\mathbf{S}} \rightarrow \mathbb{R} $ be the FIF of features $ \mathbf{X}_{\mathbf{S}} $ on $ f $. The additivity axiom states that the total influences of all possible subset of non-sensitive features is equal to the bias of the classifier.
	\begin{align}\label{eq:constraint_framework}
		f(\alg, \mathbf{D}) = \sum\nolimits_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} w_{\mathbf{S}}
	\end{align}
\end{axiom}
\noindent\textbf{Computing FIF: Reduction to Variance Decomposition.} Henceforth, our objective is to design an estimator of FIF $ w_\mathbf{S} $ that satisfies Axiom~\ref{axm:additivity}. In the following, we discuss the closed form representation of $ w_{\mathbf{S}} $ for statistical parity fairness metric and extend to other metrics in Section~\ref{sec:fairxplainer}.

@@

\noindent\textbf{Connecting Statistical Parity and Conditional Variance.} Let $ \mathbf{a}_{\max} = \argmax_{\mathbf{a}} \Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}] $ and $ \mathbf{a}_{\min} = \argmin_{\mathbf{a}} \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}] $ be the most favored and the least favored sensitive group of the classifier, respectively, according to their conditional probability of positive prediction. Next, we consider a Bernoulli random variable $ B_{\mathbf{a}} $ to denote the positive prediction of the classifier for the sensitive group $ \sensitive = \mathbf{a} $. Then, the probability of the random variable is $ p_\mathbf{a} \triangleq \Pr[B_{\mathbf{a}} = 1] = \Pr[\widehat{Y} = 1| \sensitive = \mathbf{a}] $, and variance is $ V_{\mathbf{a}} = p_{\mathbf{a}}(1-p_{\mathbf{a}}) $. Therefore, considering two random variables $ B_{\mathbf{a}_{\max}} $ and $ B_{\mathbf{a}_{\min}} $, we state the statistical parity of the classifier as $ p_{a_{\max}} - p_{a_{\min}} = (V_{a_{\max}} - V_{a_{\min}})/(1  - (p_{a_{\max}} + p_{a_{\min}})) $, where $ p_{a_{\max}} $ and $ p_{a_{\min}} $ is the probability of positive prediction of the classifier for the group $ \sensitive = \mathbf{a}_{\max} $ and $ \sensitive = \mathbf{a}_{\min} $, respectively. Intuitively, the statistical parity of the classifier is the scaled (by $ 1/(1  - (p_{a_{\max}} + p_{a_{\min}})) $) difference of the conditional variance of the positive prediction of the classifier over the most and the least favored sensitive groups.

\begin{example}
	In Example~\ref{ex:motivating_example} (Figure~\ref{fig:dt_original}), the conditional probabilities of positive prediction of the decision tree for the most and least favored groups are $  p_{a_{\max}} = \Pr[\widehat{Y} = 1 | \text{age } < 40] = 0.704 $ and $ p_{a_{\min}} = \Pr[\widehat{Y} = 1 | \text{age } \ge 40] = 0.172 $, respectively. Thus, the statistical parity of the classifier is $ p_{a_{\max}} - p_{a_{\min}} = 0.704 - 0.172 =  0.532 $. Now, we compute conditional variances of positive prediction as $ V_{a_{\max}} = \mathsf{Var}[\widehat{Y} = 1| \text{age } < 40] = 0.208 $ and $ V_{a_{\min}} = \mathsf{Var}[\widehat{Y} = 1| \text{age } \ge 40] = 0.142 $. Thus, following our argument, we compute the scaled difference in conditional variances as $ (0.208 - 0.142)/(1 - (0.702 + 0.172)) = 0.532 $, which coincides with the statistical parity.
\end{example}
Now, we apply variance decomposition (Eq.~\eqref{eq:variance_decomposition_set_notation}) for both the conditional variances $ V_{\mathbf{a}_{\max}} $ and $ V_{\mathbf{a}_{\min}} $ to compute the FIF of features, as stated in Theorem~\ref{lemma:fif}.
\begin{theorem}[FIF as Difference of Conditional Variances]
	\label{lemma:fif}
	Let $ V_{\mathbf{a}, \mathbf{S}} \triangleq \mathsf{Var}_{\nonsensitive_{\mathbf{S}}}[\widehat{Y} = 1|\sensitive = \mathbf{a}] $ be the decomposed conditional variance of the classifier's positive prediction for the sensitive group $ \sensitive = \mathbf{a} $, where features $ \nonsensitive_{\mathbf{S}} $ are jointly varied. Then, we can express FIF of $ \nonsensitive_{\mathbf{S}} $ as
	\begin{equation}\label{eq:fif_decompose}
	w_{\mathbf{S}}  = \frac{V_{\mathbf{a}_{\max},\mathbf{S}} - V_{\mathbf{a}_{\min},\mathbf{S}}}{1 - (\Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}_{\max}] + \Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}_{\min}])}
	\end{equation}
	and FIFs defined by Equation~\eqref{eq:fif_decompose} satisfy Axiom~\ref{axm:additivity}.
\end{theorem}
For brevity of space. we defer the detailed proof to Appendix~\ref{sec:proof_1}.

\noindent\textbf{Consequences of Theorem~\ref{lemma:fif}.} Here, we discuss the algorithmic consequences of Theorem~\ref{lemma:fif} and probable issues that might appear in computation.

>>>>

%in Section~\ref{??}, and conclude by extending influence computation to other group fairness metrics in Section~\ref{??}
%\subsection{An Axiomatic Formulation}
%\label{sec:problem_formulation}
\paragraph{FIF: An Axiomatic Formulation.} We are given a binary classifier  $\alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} $, a dataset $ \mathbf{D} = \{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n $, and a  fairness metric $ f(\alg, \mathbf{D}) \in \mathbb{R}^{\geq 0} $. Our objective is to compute the influences of non-sensitive features on $ f $. We compute influence of each subset of non-sensitive features $ \mathbf{X}_{\mathbf{S}} $, where $ \mathbf{S} = \{S_i | 1 \le S_i \le \numnonsensitive\} \subseteq [\numnonsensitive] \setminus \emptyset $ is a non-empty subset of indices.

\begin{definition}[Fairness Influence Function]
Fairness Influence Function (FIF) $ w_{\mathbf{S}}:\mathbf{X}_{\mathbf{S}} \rightarrow \mathbb{R} $ measures the quantitative contribution of the subset of features $ \mathbf{X}_{\mathbf{S}} \subseteq \mathbf{X}_{[k]}$ on the incurred bias $ f(\alg, \mathbf{D}) $ of the classifier $ \alg $ for dataset $\mathbf{D}$. We refer $ w_{\mathbf{S}} $ as an \emph{individual influence} when $ |\mathbf{S}| = 1 $ and an \emph{inter-sectional influence} when $ |\mathbf{S}| > 1 $\textemdash particularly, a second-order influence when $ |\mathbf{S}| = 2 $ and a higher-order influence when $ 3 \le |\mathbf{S}| \le \numnonsensitive $. %\unsure{What do we mean by function here? Isn't FIF scalar?}
\end{definition}	
Since FIF is meant to attribute the bias of the classifier to the subset of features, it is natural to choose a function such that the FIFs of all the subsets of features exactly adds up to the total bias. This leads to the following additivity axiom.
%{\fairXplainer} is an axiomatic approach to relate individual and inter-sectional influences with associated fairness metrics, which we formalize in the following.
\begin{axiom}[Additivity of influence]
	\label{axm:additivity}
	Let $ f(\alg, \mathbf{D}) \in \mathbb{R}^{\geq 0}$ be the bias of the classifer and $ w_{\mathbf{S}}:\mathbf{X}_{\mathbf{S}} \rightarrow \mathbb{R} $ be the FIF of features $ \mathbf{X}_{\mathbf{S}} $ on $ f $. The additivity axiom states that the total influences of all possible subset of non-sensitive features is equal to the bias of the classifier.
	\begin{align}\label{eq:constraint_framework}
		f(\alg, \mathbf{D}) = \sum\nolimits_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} w_{\mathbf{S}}
	\end{align}
\end{axiom}
\paragraph{Computing FIF: Reduction to Variance Decomposition.} Henceforth, our objective is to design an estimator of FIF $ w_\mathbf{S} $ that satisfies Axiom~\ref{axm:additivity}. In the following, we discuss the closed form representation of $ w_{\mathbf{S}} $ for statistical parity fairness metric and extend to other metrics in Section~\ref{sec:fairxplainer}.

@@

\paragraph{Connecting Statistical Parity and Conditional Variance.} Let $ \mathbf{a}_{\max} = \argmax_{\mathbf{a}} \Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}] $ and $ \mathbf{a}_{\min} = \argmin_{\mathbf{a}} \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}] $ be the most favored and the least favored sensitive group of the classifier, respectively, according to their conditional probability of positive prediction. Next, we consider a Bernoulli random variable $ B_{\mathbf{a}} $ to denote the positive prediction of the classifier for the sensitive group $ \sensitive = \mathbf{a} $. Then, the probability of the random variable is $ p_\mathbf{a} \triangleq \Pr[B_{\mathbf{a}} = 1] = \Pr[\widehat{Y} = 1| \sensitive = \mathbf{a}] $, and variance is $ V_{\mathbf{a}} = p_{\mathbf{a}}(1-p_{\mathbf{a}}) $. Therefore, considering two random variables $ B_{\mathbf{a}_{\max}} $ and $ B_{\mathbf{a}_{\min}} $, we state the statistical parity of the classifier as $ p_{a_{\max}} - p_{a_{\min}} = (V_{a_{\max}} - V_{a_{\min}})/(1  - (p_{a_{\max}} + p_{a_{\min}})) $, where $ p_{a_{\max}} $ and $ p_{a_{\min}} $ is the probability of positive prediction of the classifier for the group $ \sensitive = \mathbf{a}_{\max} $ and $ \sensitive = \mathbf{a}_{\min} $, respectively. Intuitively, the statistical parity of the classifier is the scaled (by $ 1/(1  - (p_{a_{\max}} + p_{a_{\min}})) $) difference of the conditional variance of the positive prediction of the classifier over the most and the least favored sensitive groups.

\begin{example}
	\normalfont
	In Example~\ref{ex:motivating_example} (Figure~\ref{fig:dt_original}), the conditional probabilities of positive prediction of the decision tree for the most and least favored groups are $  p_{a_{\max}} = \Pr[\widehat{Y} = 1 | \text{age } < 40] = 0.704 $ and $ p_{a_{\min}} = \Pr[\widehat{Y} = 1 | \text{age } \ge 40] = 0.172 $, respectively. Thus, the statistical parity of the classifier is $ p_{a_{\max}} - p_{a_{\min}} = 0.704 - 0.172 =  0.532 $. Now, we compute conditional variances of positive prediction as $ V_{a_{\max}} = \mathsf{Var}[\widehat{Y} = 1| \text{age } < 40] = 0.208 $ and $ V_{a_{\min}} = \mathsf{Var}[\widehat{Y} = 1| \text{age } \ge 40] = 0.142 $. Thus, following our argument, we compute the scaled difference in conditional variances as $ (0.208 - 0.142)/(1 - (0.702 + 0.172)) = 0.532 $, which coincides with the statistical parity.
\end{example}
Now, we apply variance decomposition (Eq.~\eqref{eq:variance_decomposition_set_notation}) for both the conditional variances $ V_{\mathbf{a}_{\max}} $ and $ V_{\mathbf{a}_{\min}} $ to compute the FIF of features, as stated in Theorem~\ref{lemma:fif}.
\begin{theorem}[FIF as Difference of Conditional Variances]
	\label{lemma:fif}
	Let $ V_{\mathbf{a}, \mathbf{S}} \triangleq \mathsf{Var}_{\nonsensitive_{\mathbf{S}}}[\widehat{Y} = 1|\sensitive = \mathbf{a}] $ be the decomposed conditional variance of the classifier's positive prediction for the sensitive group $ \sensitive = \mathbf{a} $, where features $ \nonsensitive_{\mathbf{S}} $ are jointly varied. Then, we can express FIF of $ \nonsensitive_{\mathbf{S}} $ as
	\begin{equation}\label{eq:fif_decompose}
	w_{\mathbf{S}}  = \frac{V_{\mathbf{a}_{\max},\mathbf{S}} - V_{\mathbf{a}_{\min},\mathbf{S}}}{1 - (\Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}_{\max}] + \Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}_{\min}])}
	\end{equation}
	and FIFs defined by Equation~\eqref{eq:fif_decompose} satisfy Axiom~\ref{axm:additivity}.
\end{theorem}

\begin{proof}

	Let $ B_{\mathbf{a}} \in \{0, 1\} $ be a Bernoulli random variable such that $ B_{\mathbf{a}} = 1 $ denotes the classifier predicting positive class $ \hat{Y} = 1 $ for an individual belonging to the sensitive group $ \sensitive = \mathbf{a} $, and $ B_{\mathbf{a}} = 0 $ denotes $ \hat{Y} = 0 $. Hence, if $ p_\mathbf{a} \triangleq \Pr[B_\mathbf{a} = 1] $, then $ p_\mathbf{a} $ is also the conditional probability of positive prediction of the classifier,  $ p_\mathbf{a} = \Pr[\hat{Y} = 1 |  \sensitive = \mathbf{a}] $. The variance of $ B_{\mathbf{a}} $ is computed as
	\begin{align*}
	\mathsf{Var}[B_\mathbf{a}] = p_{\mathbf{a}}(1 - p_{\mathbf{a}}). 
	\end{align*}

	Now, we consider two Bernoulli random variables $ B_\mathbf{a} $ and $ B_\mathbf{a'} $ associated with two different sensitive groups $ \sensitive = \mathbf{a} $ and $ \sensitive = \mathbf{a}' $, respectively. The difference in variances between $ B_\mathbf{a} $ and $ B_\mathbf{a'} $ implies that
	\begin{align*}
	\mathsf{Var}[B_\mathbf{a}] - \mathsf{Var}[B_\mathbf{a'}] &= p_{\mathbf{a}}(1 - p_{\mathbf{a}}) - p_{\mathbf{a'}}(1 - p_{\mathbf{a'}}) \\
	&= p_{\mathbf{a}} - p_{\mathbf{a}}^2 - p_{\mathbf{a'}} + p_{\mathbf{a'}}^2 \\ 
	&= (p_\mathbf{a} -  p_\mathbf{a'}) (1 - (p_\mathbf{a} + p_\mathbf{a'}))
	\end{align*}

	After reorganization, we express the difference in probabilities $ p_\mathbf{a} -  p_\mathbf{a'} $ as
	\begin{align}
	\label{eq:sp_to_var}
	p_\mathbf{a} -  p_\mathbf{a'}  = \frac{	\mathsf{Var}[B_\mathbf{a}] - \mathsf{Var}[B_\mathbf{a'}]}{1 - (p_\mathbf{a} + p_\mathbf{a'})}
	\end{align}

	By setting $ \mathbf{a} = \mathbf{a}_{\max} $ and $ \mathbf{a'} = \mathbf{a}_{\min} $ in Equation~\eqref{eq:sp_to_var}, we express the statistical parity of the classifier in terms of the scaled difference in the conditional variance of the positive prediction of the classifier. 
	\begin{align}
	f_{\mathsf{SP}}(\alg, \mathbf{D}) \triangleq p_{\mathbf{a}_{\max}} -  ~~p_{\mathbf{a}_{\min}}  &= \frac{	\mathsf{Var}[B_{\mathbf{a}_{\max}}] - \mathsf{Var}[B_{\mathbf{a}_{\min}}]}{1 - (p_{\mathbf{a}_{\max}} + p_{\mathbf{a}_{\min}})}\label{eq:sp_var_diff}\\
	&= \frac{	\mathsf{Var}[\hat{Y} = 1| \sensitive = \mathbf{a}_{\max}] - \mathsf{Var}[\hat{Y} = 1| \sensitive = \mathbf{a}_{\min}]}{1 - (p_{\mathbf{a}_{\max}} + p_{\mathbf{a}_{\min}})} \notag.
	\end{align}

	%The last inequality is obtained by the definition of $B_{\mathbf{a}}$.

	Next, we apply variance decomposition (Equation~\eqref{eq:variance_decomposition_set_notation}) to estimate the influence of each $ \nonsensitive_{\mathbf{S}} $. From Equation~\eqref{eq:variance_decomposition_set_notation}, we obtain that the variance of $ B_{\mathbf{a}} $ can be decomposed as 
	\[ \mathsf{Var}[B_\mathbf{a}] = \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} V_{\mathbf{a}, \mathbf{S}},\] 
	where $ V_{\mathbf{a}, \mathbf{S}} $ denotes the decomposed variance of $ B_\mathbf{a} $ w.r.t.\ the features $ \nonsensitive_{\mathbf{S}} $ conditioned on the sensitive group $ \sensitive = \mathbf{a} $.

	Now, we apply variance decomposition on both $ \mathsf{Var}[B_{\mathbf{a}_{\max}}] $ and $ \mathsf{Var}[B_{\mathbf{a}_{\min}}] $ in Equation~\ref{eq:sp_var_diff} to obtain

	\begin{align}
	f_{\mathsf{SP}}(\alg, \mathbf{D}) & = \frac{\mathsf{Var}[B_{\mathbf{a}_{\max}}] - \mathsf{Var}[B_{\mathbf{a}_{\min}}]}{1 - (p_{\mathbf{a}_{\max}} + p_{\mathbf{a}_{\max}})}= \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} \frac{V_{\mathbf{a}_{\max}, \mathbf{S}} - V_{\mathbf{a}_{\min}, \mathbf{S}}}{1 - (p_{\mathbf{a}_{\max}} + p_{\mathbf{a}_{\min}})} \label{eq:influence_value}
	\end{align}
	Finally, we estimate  the FIF of the subset of features $ \nonsensitive_{\mathbf{S}}  $ as 
	\begin{align}
	w_{\mathbf{S}}  = \frac{V_{\mathbf{a}_{\max}, \mathbf{S}} - V_{\mathbf{a}_{\min}, \mathbf{S}}}{1 - (p_{\mathbf{a}_{\max}} + p_{\mathbf{a}_{\min}})}
	\end{align} 
	by separating the terms corresponding to $ \nonsensitive_{\mathbf{S}} $.

	Following Equation~\eqref{eq:influence_value}, $	w_{\mathbf{S}} $ satisfies Axiom~\ref{axm:additivity} as $f_{\mathsf{SP}}(\alg, \mathbf{D}) = \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} w_{\mathbf{S}} $.
\end{proof}

\paragraph{Consequences of Theorem~\ref{lemma:fif}.} Here, we discuss the algorithmic consequences of Theorem~\ref{lemma:fif} and probable issues that might appear in computation.

### 2022-06-02 chapters/fairness/fif/introduction.tex
all papers added
minor edits in many parts of the thesis

\section{Introduction}
%\begin{itemize}
%	\item Bias -> fairness metric 
%	\item Fairness influence function (FIF)
%	\item Example
%	\item Contribution (1) framework (2) algorithm (3) experimental validation
%	\item Related works.
%	\item Notation paragraph
%\end{itemize}
The last decades have witnessed a significant progress in machine learning with applications in high-stake decision making, such as college admission~\cite{martinez2021using}, recidivism prediction~\cite{tollenaar2013method}, job applications~\cite{ajunwa2016hiring} etc. In such applications, the deployed machine learning classifier often demonstrate bias towards certain demographic groups involved in the data~\cite{dwork2012fairness}. For example, a classifier deciding the eligibility of college admission may offer more admission to White-male candidates than to Black-female candidates\textemdash possibly because of the historical bias in the admission data, or the accuracy-centric learning objective of the classifier, or a combination of both~\cite{berk2019accuracy,landy1978correlates,zliobaite2015relation}. Following such phenomena, multiple fairness metrics, such as \textit{statistical parity}, \textit{equalized odds}, \textit{predictive parity} etc, have been proposed to quantify the bias of the classifier on a dataset. For example, if the classifier in college admission demonstrates a {statistical parity} of $ 0.6 $, it means that White-male candidates are offered admission $ 60\% $ more than Black-female candidates~\cite{besse2021survey,feldman2015certifying,garg2020fairness}.

@@

\noindent\textbf{Our Contributions.} The contribution of this paper is three-fold.

@@

We illustrate the usefulness of our contributions via an example scenario proposed in~\cite{ghosh2020justicia}.

@@

 %\red{This allows practitioners to deploy various affirmative or punitive actions.}\improvement{Examples?}
\begin{example}
	\label{ex:motivating_example}
 Following~\cite{ghosh2020justicia}, we consider a classifier that decides an individual's eligibility for health insurance based on non-sensitive features `fitness' and `income'.
	`fitness' and `income' depend on a sensitive feature `age' leading to two sensitive groups ``young" (age $ < 40 $) and ``elderly" (age $ \ge 40 $), as highlighted in (Figure~\ref{fig:dag_age_income_fitness}--\ref{fig:distribution_example}).

@@

	Applying techniques developed in this paper, we  investigate the source of unfairness of DT$ 1 $ by computing FIFs of features, where \emph{positive numbers denote a reduction in fairness and negative numbers denotes fairness improvement}. In Figure~\ref{fig:fif_original}, fitness $(\mathrm{FIF} = 0.74)$, and the joint effect of fitness and income $(\mathrm{FIF}  = 0.05)$ cause higher statistical parity, i.e. higher bias. This observation is invoked as DT$ 1 $ predicts positively for the higher values of fitness (root node in DT$ 1 $) and elderly individuals often possess lower fitness (Figure~\ref{fig:distribution_example}). In contrast, the income of individuals  $(\mathrm{FIF}  = -0.33)$ decreases statistical parity, as elderly individuals have a higher income but lower fitness.

	\textbf{Case study 2:} Since DT$ 1 $ is unfair to the ``elderly", we learn another decision tree (DT$ 2 $) by applying an affirmative action, where we decrease the threshold on income from $ 0.69 $ to $ 0.555 $ when the fitness is lower ({\color{affirmative}green} node in Figure~\ref{fig:dt_affirmative}). This action allows more elderly individuals to receive insurance by including more people with lower fitness, and thus the statistical parity becomes $ \Pr[\widehat{Y} = 1 | \text{age } < 40] - \Pr[\widehat{Y} = 1 | \text{age } \ge 40] = 0.71 - 0.7 = 0.01$. This is significantly less than the earlier statistical parity of $ 0.53 $.	Again, applying techniques developed in this paper, we compute FIF of features in Figure~\ref{fig:fif_affirmative}. Here, the FIF of income and fitness reflects the reduction in statistical parity as their influences almost nullify each other. Thus, FIF depicts the effects of different features and their combinations on the resultant bias incurred by the classifier. %As we analyse the source of unfairness, the combined influence of income and fitness changes its direction ($ - 0.01 $ vs.\ $ 0.05 $) towards improving fairness.
\end{example}

@@

\section{Related Work} Recently, several studies apply \emph{local explanation methods} of black-box prediction to explain sources of bias by feature-attribution~\cite{begley2020explainability,lundberg2020explaining} and causal path decomposition~\cite{pan2021explaining}. Since our work adopts feature-attribution approach, it reveals two-fold limitations of existing methods: (i) \emph{inaccuracy} in computing FIFs and (ii) \emph{failing to compute intersectional} FIFs. Elaborately, FIF computation in~\cite{begley2020explainability,lundberg2020explaining} is inaccurate because of applying local explainers such as SHAP~\cite{lundberg2017unified} to compute global properties of the classifier such as group fairness. In addition, features are often correlated in practical fairness tasks, and computing only individual FIFs ignores the joint contribution of multiple features on the unfairness of the classifer. Also, these works provide empirical evaluations to justify the choices of SHAP-based tools for explaining fairness and does not consider the global nature of group fairness metrics. In this paper, \textit{we develop a formal framework to explain sources of unfairness in a classifier and also a novel methodology to address it. To the best of our knowledge, this is the first work to do the both}. %\todo{1. GSA with fairness, 2. Citation not correct, 3. Influence of Yair Zick.}
Among other related works, \cite{benesse2021fairness} link GSA measures such as Sobol and Cram{\'e}r-von-Mises indices to different fairness metrics. While their approach relates the GSA of sensitive features on the resulting bias, we focus on applying GSA to all features to compute FIFs. \textit{Their approach only detects the presence or absence of bias, while we focus on decomposing bias as the sum of FIFs of all features}. In another line of work, \cite{datta2016algorithmic} and \cite{ghosh2022algorithmic} compute feature-influence as the shift of bias from its original value by randomly intervening features. Their work is different from our axiomatic approach, where the sum of FIFs equals the bias.

>>>>

%\section{Introduction}
%\begin{itemize}
%	\item Bias -> fairness metric 
%	\item Fairness influence function (FIF)
%	\item Example
%	\item Contribution (1) framework (2) algorithm (3) experimental validation
%	\item Related works.
%	\item Notation paragraph
%\end{itemize}
The last decades have witnessed a significant progress in machine learning with applications in high-stake decision making, such as college admission~\cite{martinez2021using}, recidivism prediction~\cite{tollenaar2013method}, job applications~\cite{ajunwa2016hiring} etc. In such applications, the deployed machine learning classifier often demonstrate bias towards certain demographic groups involved in the data~\cite{dwork2012fairness}. For example, a classifier deciding the eligibility of college admission may offer more admission to White-male candidates than to Black-female candidates\textemdash possibly because of the historical bias in the admission data, or the accuracy-centric learning objective of the classifier, or a combination of both~\cite{berk2019accuracy,landy1978correlates,zliobaite2015relation}. Following such phenomena, multiple fairness metrics, such as \textit{statistical parity}, \textit{equalized odds}, \textit{predictive parity} etc, have been proposed to quantify the bias of the classifier on a dataset. For example, if the classifier in college admission demonstrates a {statistical parity} of $ 0.6 $, it means that White-male candidates are offered admission $ 60\% $ more than Black-female candidates~\cite{besse2021survey,feldman2015certifying,garg2020fairness}.

@@

\paragraph{Our Contributions.} The contribution of this chapter is three-fold.

@@

We illustrate the usefulness of our contributions via an example scenario in Example~\ref{fairness_justicia_example:intro}.

@@

 %\red{This allows practitioners to deploy various affirmative or punitive actions.}\improvement{Examples?}
\begin{example}
	\normalfont
	\label{ex:motivating_example}
 Following~\cite{ghosh2020justicia}, we consider a classifier that decides an individual's eligibility for health insurance based on non-sensitive features `fitness' and `income'.
	`fitness' and `income' depend on a sensitive feature `age' leading to two sensitive groups ``young" (age $ < 40 $) and ``elderly" (age $ \ge 40 $), as highlighted in (Figure~\ref{fig:dag_age_income_fitness}--\ref{fig:distribution_example}).

@@

	Applying techniques developed in this chapter, we  investigate the source of unfairness of DT$ 1 $ by computing FIFs of features, where \emph{positive numbers denote a reduction in fairness and negative numbers denotes fairness improvement}. In Figure~\ref{fig:fif_original}, fitness $(\mathrm{FIF} = 0.74)$, and the joint effect of fitness and income $(\mathrm{FIF}  = 0.05)$ cause higher statistical parity, i.e. higher bias. This observation is invoked as DT$ 1 $ predicts positively for the higher values of fitness (root node in DT$ 1 $) and elderly individuals often possess lower fitness (Figure~\ref{fig:distribution_example}). In contrast, the income of individuals  $(\mathrm{FIF}  = -0.33)$ decreases statistical parity, as elderly individuals have a higher income but lower fitness.

	\textbf{Case study 2:} Since DT$ 1 $ is unfair to the ``elderly", we learn another decision tree (DT$ 2 $) by applying an affirmative action, where we decrease the threshold on income from $ 0.69 $ to $ 0.555 $ when the fitness is lower ({\color{affirmative}green} node in Figure~\ref{fig:dt_affirmative}). This action allows more elderly individuals to receive insurance by including more people with lower fitness, and thus the statistical parity becomes $ \Pr[\widehat{Y} = 1 | \text{age } < 40] - \Pr[\widehat{Y} = 1 | \text{age } \ge 40] = 0.71 - 0.7 = 0.01$. This is significantly less than the earlier statistical parity of $ 0.53 $.	Again, applying techniques developed in this chapter, we compute FIF of features in Figure~\ref{fig:fif_affirmative}. Here, the FIF of income and fitness reflects the reduction in statistical parity as their influences almost nullify each other. Thus, FIF depicts the effects of different features and their combinations on the resultant bias incurred by the classifier. %As we analyse the source of unfairness, the combined influence of income and fitness changes its direction ($ - 0.01 $ vs.\ $ 0.05 $) towards improving fairness.
\end{example}

@@

\section{Related Work} Recently, several studies apply \emph{local explanation methods} of black-box prediction to explain sources of bias by feature-attribution~\cite{begley2020explainability,lundberg2020explaining} and causal path decomposition~\cite{pan2021explaining}. Since our work adopts feature-attribution approach, it reveals two-fold limitations of existing methods: (i) \emph{inaccuracy} in computing FIFs and (ii) \emph{failing to compute intersectional} FIFs. Elaborately, FIF computation in~\cite{begley2020explainability,lundberg2020explaining} is inaccurate because of applying local explainers such as SHAP~\cite{lundberg2017unified} to compute global properties of the classifier such as group fairness. In addition, features are often correlated in practical fairness tasks, and computing only individual FIFs ignores the joint contribution of multiple features on the unfairness of the classifer. Also, these works provide empirical evaluations to justify the choices of SHAP-based tools for explaining fairness and does not consider the global nature of group fairness metrics. In this chapter, \textit{we develop a formal framework to explain sources of unfairness in a classifier and also a novel methodology to address it. To the best of our knowledge, this is the first work to do the both}. %\todo{1. GSA with fairness, 2. Citation not correct, 3. Influence of Yair Zick.}
Among other related works, \cite{benesse2021fairness} link GSA measures such as Sobol and Cram{\'e}r-von-Mises indices to different fairness metrics. While their approach relates the GSA of sensitive features on the resulting bias, we focus on applying GSA to all features to compute FIFs. \textit{Their approach only detects the presence or absence of bias, while we focus on decomposing bias as the sum of FIFs of all features}. In another line of work, \cite{datta2016algorithmic} and \cite{ghosh2022algorithmic} compute feature-influence as the shift of bias from its original value by randomly intervening features. Their work is different from our axiomatic approach, where the sum of FIFs equals the bias.

### 2022-06-02 chapters/fairness/fvgm/conclusion.tex
minor edits in many parts of the thesis

\section{Chapter Summary}
In this paper, we propose {\fvgm}, an efficient fairness verification framework for linear classifiers based on a novel stochastic subset-sum problem. {\fvgm} encodes a graphical model of feature-correlations, represented as a Bayesian Network, and computes multiple group and causal fairness metrics accurately. We experimentally demonstrate that {\fvgm} is not only more accurate and scalable than the existing verifiers but also applicable in practical fairness tasks, such as verifying fairness attacks and enhancing algorithms, and computing the fairness influence functions. 
As a future work, we aim to design fairness-enhancing algorithms certified by fairness verifiers, such as {\fvgm}. %We also investigate a principled approach to detect the sources of bias by computing fairness influence functions (FIF). 
Since {\fvgm} serves as an accurate and scalable fairness verifier for linear classifiers, it will be interesting to design such verifiers for other ML models.
%Though the proposed solution with a Bayesian network has exponential worst-case complexity, it would be interesting to develop a pseudo-polynomial approximation in the future.

>>>>

\section{Chapter Summary}
In this chapter, we propose {\fvgm}, an efficient fairness verification framework for linear classifiers based on a novel stochastic subset-sum problem. {\fvgm} encodes a graphical model of feature-correlations, represented as a Bayesian Network, and computes multiple group and causal fairness metrics accurately. We experimentally demonstrate that {\fvgm} is not only more accurate and scalable than the existing verifiers but also applicable in practical fairness tasks, such as verifying fairness attacks and enhancing algorithms, and computing the fairness influence functions. 
As a future work, we aim to design fairness-enhancing algorithms certified by fairness verifiers, such as {\fvgm}. %We also investigate a principled approach to detect the sources of bias by computing fairness influence functions (FIF). 
Since {\fvgm} serves as an accurate and scalable fairness verifier for linear classifiers, it will be interesting to design such verifiers for other ML models.
%Though the proposed solution with a Bayesian network has exponential worst-case complexity, it would be interesting to develop a pseudo-polynomial approximation in the future.
